Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-03-31 tagless/tagless-final.tex
A bit more work on introduction

\Citet{WalidICFP02} and \citet{xi-guarded} motivated interpreting
a typed object language in a typed metalanguage as an interesting
problem.  The known solutions to this problem store object terms and
values in the metalanguage in a universal type, a generalized algebraic
datatype (GADT), or a dependent type.  In the remainder of this section,
we discuss these solutions, identify their drawbacks, then summarize our
proposal and contributions.  No matter how we represent the object language in the
metalanguage, the representation can be performed by hand or more
conveniently by a parser or preprocessor such as camlp4.  We leave aside
the issue of whether to program a parser or preprocessor using dependent
types \citep{WalidICFP02} or not \citep{baars-typing}.

@@

The unfortunate tag |UB| in the result reflects that |eval| is a partial
function, in two ways.  First, the pattern match |with UA f| in the line
|A (e1,e2)| is not exhaustive, so |eval| can fail if we apply a boolean,
as in the ill-typed term |A (B true, B false)|.
\begin{code}
let test2 = A (B true, B false)
let test2r = eval [] test2
Exception: Match_failure in eval
\end{code}
Second, the |lookup|
function assumes a nonempty environment, so |eval| can fail if we
evaluate an open term.
\begin{code}
let test3 = V VZ
let test3r = eval [] test3
Exception: Match_failure in lookup
\end{code}
After all, the type |exp| represents object
terms both well-typed and ill-typed, both open and closed.

If we evaluate only closed terms that have been type-checked, then
|eval| would never fail. Alas, this soundness is not obvious to the
metalanguage, whose type system we must still appease with the
nonexhaustive pattern matching in |lookup| and |eval| and the tags |UB|
and |UA| \citep[\S1.4]{WalidICFP02}.  In other words, the algebraic data
types above fail to express in the metalanguage that the object program
is well-typed.  This failure necessitates tagging and nonexhaustive
pattern\hyp matching operations that incur a performance penalty in
interpretation and impair optimality in partial evaluation
\citep{taha-tag}.  In short, the universal\hyp type solution is
unsatisfactory because it does not preserve typing.

@@

If we change the encoding of the term from data
constructors to functions, we can encode a tagless interpreter for a
typed object language into a metalanguage with a very simple type system.

@@

We have programmed our interpreters in OCaml (and, for staging,
MetaOCaml (\cite{metaocaml})) and Haskell. The complete code is available 
\footnote{the URL will appear in the final version of the paper.}
% connot give URL in the ICFP paper due to double-blind reviews
in the supplemental material to the paper. We will use both
(Meta)OCaml and Haskell for our examples throughout the paper.
Each language has its strength and weaknesses; this dual implementation
forced us to clarify our own ideas.  We will continually leverage the
respective strengths of these languages, as certain of our claims are
patently obvious in one implementation, while requiring more argumentation
in the other.
Our code is surprisingly simple, and, in the hindsight, obvious. We
must emphasize however that direct typeful tagless interpretation of a typed
language until now has thought to be impossible in plain ML and
require extensions like GADT -- which we discuss in detail in Section
XXX.

Now, let us define our object language (embedded DSL) in a different
way. Instead of using algebraic datatypes to encode the terms of the
language in the meta-language (reminiscent to the \emph{initial algebra} 
approach), we will be using functions as term constructors (the \emph{final}
approach):

\begin{code}
  let b (bv:bool) env  = bv
  let varZ env         = fst env
  let varS vp env      = vp (snd env)
  let app e1 e2 env    = (e1 env) (e2 env)
  let lam e env        = fun x -> e (x,env)
\end{code}

\noindent That is all of the interpreter. Our sample term now reads
\begin{code}
  let testf1 = app (lam varZ) (b true)
\end{code}
which is almost the same as before, only written with lower-case
identifiers. To evaluate a term, we merely need to apply 
the term to the empty environment

\begin{code}
  let testf1r = testf1 ()
  val testf1r : bool = true
\end{code}

The result now has no tags. The above functions |b|,
|varZ|, |app| constitute the evaluator. It is patent that the
interpreter has no tags and no pattern matching. The term |b true| is
evaluated to a boolean and a term |lam varZ| is evaluated to a
function, with no tags attached. The application rule applies that
function without any need for pattern-matching. The evaluator has
another feature: evaluation of open terms leads to a type error rather
than a run-time error. Indeed,

\begin{code}
  let testf1bad = app (lam (varS varZ)) (b true)
  let testf1badr = testf1bad ()
\end{code}
gives a type error saying 
that the initial environment should be a tuple rather than () --
meaning the term requires a non-empty initial environment: the term is
open.

Thus all the errors that were previously reported at run-time (open terms,
applications of non-functions) are now reported at compile
time. Running of the interpreter will never result in run-time errors
because its code uses no operations that could possibly raise such
an error. In fact, our interpreter is total.

Furthermore, it should be clear that we do not use any
pattern-matching at all, and thus cannot have any pattern-matching
errors.  This is clear not just to us, but also to the compiler.
In other words in traditional approaches one can prove a meta-theorem
that all pattern-matching is total, but this is nevertheless not
apparent to the compiler, which thus still has to generate code for
those cases.  This seriously hurts efficiency~\cite{WalidICFP02}.

In the above code, the interpreter is wired in, directly built into the
functions |b|, |app|, etc.  In the remainder of this paper, 
we explain how to abstract the interpreter,
to make it possible to ``evaluate'' the same term in many different
ways: interpretation, compilation, partial evaluation, size
computation, etc.

>>>>

\Citet{WalidICFP02} and \citet{xi-guarded} motivated interpreting
a typed object language in a typed metalanguage as an interesting
problem.  The known solutions to this problem store object terms and
values in the metalanguage in a universal type, a generalized algebraic
data type (GADT), or a dependent type.  In the remainder of this section,
we discuss these solutions, identify their drawbacks, then summarize our
proposal and contributions.  No matter how we represent the object language in the
metalanguage, the representation can be performed by hand or more
conveniently by a parser or preprocessor such as camlp4.  We leave aside
the issue of whether to program a parser or preprocessor using dependent
types \citep{WalidICFP02} or not \citep{baars-typing}.

@@

The unfortunate tag |UB| in the result reflects that |eval| is a partial
function, in two ways.  First, the pattern match |with UA f| in the line
|A (e1,e2)| is not exhaustive, so |eval| can fail if we apply a boolean,
as in the ill-typed term |A (B true, B false)|.
\begin{code}
let test2 = A (B true, B false)
let test2r = eval [] test2
Exception: Match_failure in eval
\end{code}
Second, the |lookup|
function assumes a nonempty environment, so |eval| can fail if we
evaluate an open term.
\begin{code}
let test3 = A (L (V (VS VZ)), B true)
let test3r = eval [] test3
Exception: Match_failure in lookup
\end{code}
After all, the type |exp| represents object
terms both well-typed and ill-typed, both open and closed.

If we evaluate only closed terms that have been type-checked, then
|eval| would never fail. Alas, this soundness is not obvious to the
metalanguage, whose type system we must still appease with the
nonexhaustive pattern matching in |lookup| and |eval| and the tags |UB|
and |UA| \citep[\S1.4]{WalidICFP02}.  In other words, the algebraic data
types above fail to express in the metalanguage that the object program
is well-typed.  This failure necessitates tagging and nonexhaustive
pattern\hyp matching operations that incur a performance penalty in
interpretation \citep{WalidICFP02} and impair optimality in partial evaluation
\citep{taha-tag}.  In short, the universal\hyp type solution is
unsatisfactory because it does not preserve typing.

@@

We represent object programs using ordinary functions rather than
data constructors.  The interpreter below provides these functions.
\begin{code}
let varZ (x,env)    = x
let varS vp (x,env) = vp env
let b (bv:bool) env = bv
let lam e env       = fun x -> e (x,env)
let app e1 e2 env   = (e1 env) (e2 env)
\end{code}
These functions comprise the entire interpreter.
We now represent our sample term $(\fun{x}x)\True$ as
\begin{code}
let testf1 = app (lam varZ) (b true)
\end{code}
This representation is almost the same as in \S\ref{tagproblem}, only
written with lowercase identifiers. To evaluate an object term is to
apply its representation to the empty environment.
\begin{code}
let testf1r = testf1 ()
val testf1r : bool = true
\end{code}
The result has no tags---the interpreter patently uses no tags and no
pattern matching. The term |b true| evaluates to a boolean and the term
|lam varZ| evaluates to a function, both untagged. The |app| function
applies |lam varZ| without pattern matching. What is more, evaluating an
open term such as |testf3| below gives a type error rather than
a run-time error.
\begin{code}
let testf3 = app (lam (varS varZ)) (b true)
let testf3r = testf3 ()
\end{code}
The type error correctly complains
that the initial environment should be a tuple rather than |()|.
In other words, the term is open.

In sum, by encoding terms using ordinary functions (the \emph{final}
rather than \emph{initial} approach), we achieve a tagless interpreter
for a typed object language, in a metalanguage with a very simple type
system \citep{hindley-principal,milner-theory}.  All the run-time errors
in \S\ref{tagproblem} (applying a nonfunction; evaluating an open term)
are now reported at compile time. Because this new interpreter uses no
pattern matching, it never results in a run-time error and is in fact
total.  Because this totality is obvious not just to us but also to the
metalanguage implementation, we avoid the serious performance penalty
\citep{WalidICFP02} of error checking.

The interpreter above is wired directly into the
functions |b|, |lam|, and so on.  In the remainder of this paper, 
we explain how to abstract the interpreter so as
to ``evaluate'' the same term in many different
ways: interpretation, compilation, partial evaluation, size
computation, and so forth.

We have programmed our interpreters in OCaml (and, for staging,
MetaOCaml \citep{metaocaml} and Haskell. The complete code is available 
\footnote{the URL will appear in the final version of the paper.}
% connot give URL in the ICFP paper due to double-blind reviews
in the supplemental material to the paper. We will use both
(Meta)OCaml and Haskell for our examples throughout the paper.
Each language has its strength and weaknesses; this dual implementation
forced us to clarify our own ideas.  We will continually leverage the
respective strengths of these languages, as certain of our claims are
patently obvious in one implementation, while requiring more argumentation
in the other.
Our code is surprisingly simple, and, in the hindsight, obvious. We
must emphasize however that direct typeful tagless interpretation of a typed
language until now has thought to be impossible in plain ML and
require extensions like GADT -- which we discuss in detail in Section
XXX.

@@
